---
title: "hw 3, env s 193ds"
author: "Damian Rojas"
format: pdf
---


```{r}
#| label: set-up
#| message: false
#read packages
library(tidyverse)
library(janitor)
library(here)


#store kelp data
kelp <- read.csv("data/temp-kelp.csv")

#store personal data
walks <- read.csv("data/walks_data.csv")
```

## 1)
### a)
The appropriate tests we can run to test the strength of the relationship 
between temperature and giant kelp frond elongation rate are a Pearson's 
correlation test and a Spearman's rank correlation. 
The differences between the two tests are that a Pearson's test observes the 
strength between linear relationships while the Spearman's test observes the 
strength between monotonic relationships that do not need to be linear. 

## b) Visualizations

```{r}
#| echo: false

ggplot(kelp, mapping = aes( x = temp_c, y = kelp_elong)) +
# base ggplot, change color and size 
# of points
geom_point(color = "pink3", size = 3)+
#change x-axis to temp(celsius) 
#change y to kelp length (cm)  
  labs( x = "Temperature (Celsius)", 
        y = "Kelp Length (cm)") +
  #change theme from base
  theme_linedraw()



```


## c) Checking Assumptions / Run test

```{r}
#| label: assumptions
#| fig-height: 8
#| fig-width: 6

#create linear model for kelp data
kelp_model <- lm (kelp_elong ~ temp_c, data = kelp)

#show lm for kelp
kelp_model
#kelp elongation as function of temp
#
#plotted as 2x2 grid
 par(mfrow = c(2,2))
 #summary plots of kelp model
plot (kelp_model) 
 


```

Checked normality, homoscedasticity, and any outliers of data with summary plots. Found to be normal, linear, with no outliers affecting data. 

```{r}
#| label: test
#pearsons cor test is the method for linear models
#run test with x as kelp elongation, y as temp, method as pearson
cor.test(x = kelp$kelp_elong, y = kelp$temp_c, method = "pearson")
```


## d)
To evaluate the strength of the relationship between temperature and giant kelp frond elongation rate, I used a Pearson's correlation test. 
We found a strong negative relationship between kelp elongation and temperature (Pearson's r = -0.69, t(30) = -5.19, p < 0.001), where temperature significantly predicted kelp elongation length (R^2^ = 0.476).
